---
title: " 2024 ELLS summer school Darwinian agriculture"
subtitle: "Workshop Social evolution under mass selection or under true breeding values"
author: "Jacques David, Nicolas Salas, Peter Bourke"
date: "2024-06-20"
output:
  pdf_document: default
  html_document: default
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

```

# Introduction
Welcome to this practical session on the decomposition of phenotypic value and the effects of selection. In this session, we will explore how the phenotypic value of a plant is influenced by its genotype and by the genotypes of its neighbors, and how selection on these traits affects the next generation.

# Key theoretical Concepts

find the ref Bijma , Muir etc

## Decomposition of the Phenotypic Value
Classically, in one single macro-environment, the phenotypic value ($P$) of an individual can be decomposed into multiple components:
$$ P = G + E $$  
Where:  
- **G** represents the genetic contribution to the phenotype,  
- **E** represents the micro-environmental contribution to the phenotype.  

Let's rigorously derive the total phenotypic variance $V_P$ where the phenotypes may not independent, either because of genetic relatedness or structured environment.

Consider the phenotype $P_i$ of an individual $i$, influenced by its genetic effect ($G_i$) and its direct environmental effect ($e_i$): $P_i = G_i + e_i$

## Non-Independent Phenotypes

Consider the phenotype $P_i$ of an individual $i$, $P_i = G_i + e_i$,

where $G_i$ is the genetic effect and $e_i$ is the environmental effect.

For the sake of simplicity, let'use centered value of $P$, with  so that $E(P)=E(G)=E(e)=0$, and $V_P=E(P_i^2); V_G=E(G_i^2) ; V_e=E(e_i^2)$

The total phenotypic variance is defined as:

$V_P = E(P_i)^2$

First, we need to determine $E(P_i)$:  

$E(P) = \frac{1}{n} \sum_{i=1}^n P_i $  

where $n$ is the number of individuals.

$$\begin{split}
E(P_i-\frac{1}{n} \sum_{i=1}^n P_i )^2 &= E(P_i-\frac{1}{n} (P_i+..+P_j ))^2\\
&=E(P_i - \frac{1}{n}(P_iP_j + P_j + P_jP_i..))^2
\end{split}$$


Next, let's calculate \(E(P^2)\):

$P^2 = \left( \frac{1}{n} \sum_{i=1}^n P_i \right)^2 = \frac{1}{n^2} \sum_{i=1}^n \sum_{j=1}^n P_i P_j$

Taking the expectation:

\[ E(P^2) = \frac{1}{n^2} \sum_{i=1}^n \sum_{j=1}^n E(P_i P_j) \]

#### Step 3: Calculate \(E(P_i P_j)\)

We know:

\[ E(P_i P_j) = E((G_i + e_i)(G_j + e_j)) \]
\[ = E(G_i G_j) + E(G_i e_j) + E(e_i G_j) + E(e_i e_j) \]

Given that \(G_i\) and \(e_i\) are independent and have zero means, \(E(G_i e_j) = E(e_i G_j) = 0\). If \(e_i\) and \(e_j\) are also independent, \(E(e_i e_j) = 0\). Thus:

\[ E(P_i P_j) = E(G_i G_j) \]

#### Step 4: Total Phenotypic Variance \(V(P)\)

\[ V(P) = E(P^2) - (E(P))^2 \]

Since \(E(P) = 0\):

\[ V(P) = E(P^2) \]
\[ = \frac{1}{n^2} \sum_{i=1}^n \sum_{j=1}^n E(P_i P_j) \]
\[ = \frac{1}{n^2} \left( \sum_{i=1}^n E(P_i^2) + \sum_{i \neq j} E(P_i P_j) \right) \]

From earlier, we have:

\[ E(P_i^2) = \text{Var}(P_i) = \text{Var}(G_i) + \text{Var}(e_i) \]
\[ E(P_i P_j) = \text{Cov}(G_i, G_j) \]

Thus:

\[ V(P) = \frac{1}{n^2} \left( n (\text{Var}(G) + \text{Var}(e)) + n(n-1) \text{Cov}(G_i, G_j) \right) \]
\[ = \frac{1}{n} (\text{Var}(G) + \text{Var}(e)) + \frac{n-1}{n} \text{Cov}(G_i, G_j) \]

Given that we typically separate the genetic and environmental variances and their covariances, we can simplify this to:

\[ V(P) = \text{Var}(G) + \text{Var}(e) + \text{Cov}(G_i, G_j) \]

### Conclusion

The total phenotypic variance \(V(P)\) is:

\[ V(P) = V_G + V_e + \text{Cov}(G_i, G_j) \]

This expression correctly separates the genetic variance (\(V_G\)), the environmental variance (\(V_e\)), and the covariance between the genetic effects of different individuals (\(\text{Cov}(G_i, G_j)\)). This approach ensures that all sources of variation and their interactions within the population are accounted for.
## What happens if Phenotypes are not independent

Now write E(Pi - 1/n)

consider the covariance between two phenotypes \( P_i \) and \( P_j \):

\[ \text{Cov}(P_i, P_j) = E[P_i P_j] - E[P_i]E[P_j] \]

Given that \( E(P_i) = 0 \) and \( E(P_j) = 0 \), we have:

\[ \text{Cov}(P_i, P_j) = E[P_i P_j] \]

Let's expand \( P_i P_j \):

\[ P_i P_j = (DGE_i + e_i)(DGE_j + e_j) = DGE_i DGE_j + DGE_i e_j + e_i DGE_j + e_i e_j \]

Taking the expectation:

\[ E[P_i P_j] = E[DGE_i DGE_j] + E[DGE_i e_j] + E[e_i DGE_j] + E[e_i e_j] \]

Again, since \( DGE \) and \( e \) are independent, \( E[DGE_i e_j] = E[e_i DGE_j] = 0 \). If \( e_i \) and \( e_j \) are also independent, \( E[e_i e_j] = 0 \). Thus:

\[ E[P_i P_j] = E[DGE_i DGE_j] \]

So:

\[ \text{Cov}(P_i, P_j) = \text{Cov}(DGE_i, DGE_j) \]

### Total Phenotypic Variance

When considering the total phenotypic variance for a population, we need to include both the variance of individual phenotypes and the covariances between different phenotypes. The total phenotypic variance \( V(P) \) is given by:

\[ V(P) = \sum_i \text{Var}(P_i) + \sum_{i \neq j} 2 \cdot \text{Cov}(P_i, P_j) \]

#### Putting It All Together

1. **Variance of \( P_i \)**:
   \[ \text{Var}(P_i) = \text{Var}(DGE_i) + \text{Var}(e_i) \]

2. **Covariance between \( P_i \) and \( P_j \)**:
   \[ \text{Cov}(P_i, P_j) = \text{Cov}(DGE_i, DGE_j) \]

So, the total phenotypic variance \( V(P) \) is:

\[ V(P) = \sum_i \left( \text{Var}(DGE_i) + \text{Var}(e_i) \right) + \sum_{i \neq j} 2 \cdot \text{Cov}(DGE_i, DGE_j) \]

Given that we are taking expectations, the final formula can be written as:

\[ V(P) = N \left( \mathbb{E}[\text{Var}(DGE)] + \mathbb{E}[\text{Var}(e)] \right) + N(N-1) \cdot 2 \cdot \mathbb{E}[\text{Cov}(DGE_i, DGE_j)] \]

where \( N \) is the number of individuals.

### Conclusion

In the classic case, the total phenotypic variance \( V(P) \) includes:
- The variances of the direct genetic effects and environmental effects of individuals.
- The covariance between the direct genetic effects of different individuals, counted twice because each pairwise interaction contributes to the total variance.

This thorough approach ensures that we account for all sources of variation and their interactions within the population.


""""""""""""""""""""""""""""""""

In the case of heterogeneous populations, such as Bulks or landraces or farmers varieties, individuals of different genotypes are neighbouring each others. So, if we introduce the fact that the phenotype of a $focal$ individual, $i$, $P_{Focal_i}$, can be determined by its own genotype and by the genotype of its neighbour, $j$, $P_{Focal_i, Neighbour_j}$ can now be decomposed into a Direct Genetic Effect ($DGE$) due to $i$ and the Indirect Genetic Effect ($IGE$) from its neighbour  $j$, the model becomes:  

$$P_{Focal_i, Neighbour_j} = DGE_{Focal_i} + IGE_{Neignhbour_j} + E_{DGE_{Focal_i}} + E_{IGE_{Neignhbour_j}}$$
Where:  
- $DGE_{Focal_i}$ is a Direct Genetic Effect of $i$, i.e., the direct contribution of the individual's own genotype to its own phenotype,  
- $IGE_{Neignhbours}$ is the Indirect genetic contribution of the genotype of the neighbourc $j$  
and,  
- $E_{DGE_{Focal_i}}$ : environment effect on Direct Genetic Effect) on the focal plant  
- $E_{IGE_{Neignhbour_j}}$ :environment effect on Indirect Genetic Effect of the neighbour  

These components have their associated variances : $V_{DGE},V_{IGE}, V_{E_{DGE}}, V_{E_{IGE}}$  

Of course, any individual in an heterogeneous populations, if a individual $i$ has a $DGE_i$ it also have a $IGE_i$ than can be correlated.  

## Correlation between DGE and IGE
A correlation between $DGE_i$ and $IGE_i$ will affects the overall distribution of the phenotypic values and hence the  response to selection. For a social trait, a strong competitor, $i$, can have a strong positive $DGE_i$ value while reducing the trait value of its neighbours with a negative $IGE_i$ value.  

According to the trait, the correlation between the $DGE_i$ and their $IGE_i$ counterpart can theoretically vary from -1 to 1. This correlation will thus translate in the data into a covariance between the $DGE_i$ and $IGE_i$ values, denoted $cov(DGE:IGE)$ with  

$$r_{(DGE:IGE)}=\frac{cov(DGE:IGE)}{\sqrt{V_{DGE}V_{IGE}}}$$


## Phenotypic variance in a complex neighbourhood 

In our workshop situation, we will consider that plants are grown on a grid and any of them have 8 neighbours with which it is interacting. 

```{r, echo=FALSE}
# Create the table with a focal cell and its neighbors
table <- matrix(c("Neighbour1", "Neighbour2", "Neighbour3",
                  "Neighbour4", "Focal", "Neighbour5",
                  "Neighbour6", "Neighbour7", "Neighbour8"),
                nrow = 3, byrow = TRUE)
knitr::kable(table)

```


$$ \begin{split}
P_{i, (n1..n8)} = & DGE_{i} \,+ 
                        \\ & IGE_{n1} \, +..+ \,IGE_{n8} \,+ 
                        \\& E_{DGE_{i}} \,+ 
                        \\& E_{IGE_{n1}}+..+E_{IGE_{n8}}
\end{split}$$

To calculate the total phenotypic variance $V_P$, we may recall the usual case where $P=G+E$ when the phenotypic values can be considered as independent when the genotypes are not related and when there is no particular structured environment. In this case $V_P = V_P + $

If a mass selection is performed based on phenotypic value, the selection acts on both the individual's genotype and the genotypes of its neighbors, a concept known as the social phenotype.

##### True Breeding Value (TBV)
The true breeding value (TBV) integrates both DGE and IGE. It is an important metric for predicting the genetic value of the next generation, particularly when social interactions are considered. The TBV is influenced by the correlation between DGE and IGE and can be calculated as:
\[ TBV = DGE + \sum IGE \]
where \(\sum IGE\) represents the sum of the indirect genetic effects from all relevant neighbors.

##### Environmental Effects
Environmental effects (\(E\)) play a significant role in shaping the phenotypic value. These effects can be unique to each individual (specific environment) or shared among individuals (common environment). In our model, we consider environmental variances for both DGE and IGE.

#### Bibliographic Context
Understanding the contributions of genetic and environmental factors to phenotypic variance has been a core pursuit in quantitative genetics. The inclusion of social interactions, as captured by IGE, has been pivotal in advancing breeding programs, particularly in plants and animals with significant social structures. Studies (ref) have laid the foundation for integrating IGEs into genetic models, highlighting their impact on selection response and breeding strategies.

#### Simulation Algorithm
The simulation algorithm used in this practical session involves the following steps:

1. **Matrix Preparation**: Genetic (G) and environmental (E) variance-covariance matrices are prepared based on user inputs.
2. **Genotype and Neighborhood Generation**: Genotypes and environmental effects are generated for a grid of plants, and neighbors are assigned randomly.
3. **Phenotypic Value Calculation**: The phenotypic value for each plant is calculated as the sum of its DGE, the IGEs from its neighbors, and environmental effects.
4. **Selection**: A proportion of the top-performing plants is selected based on their phenotypic values.
5. **Next Generation Prediction**: The mean genetic values for the next generation are predicted based on the selected plants, taking into account the variances and covariance between DGE and IGE.

#### Detailed Steps in the Simulation

1. **Matrix Preparation**:
    - Calculate the covariance between DGE and IGE based on the user-provided genetic correlation.
    - Form the genetic variance-covariance matrix \(G\) and the environmental variance-covariance matrix \(E\).

    ```r
    covdgE_IGE <- input$r * (input$varG11 * input$varG22)^0.5
    G <- matrix(c(input$varG11, covdgE_IGE, covdgE_IGE, input$varG22), nrow = 2, byrow = TRUE)
    E <- matrix(c(input$varE11, 0, 0, input$varE22), nrow = 2, byrow = TRUE)
    ```

2. **Genotype and Neighborhood Generation**:
    - Generate genotype values using a multivariate normal distribution for a grid of \(N \times N\) plants.
    - Generate environmental effects similarly.
    - Assign neighbors randomly on a grid using a neighborhood matrix.

    ```r
    genotype <- mvrnorm(n = N^2, mu = c(0, 0), Sigma = G)
    enviro <- mvrnorm(n = N^2, mu = c(0, 0), Sigma = E)
    voisinage <- matrix(sample(1:N^2, 4*N^2, replace = TRUE), nrow = 2*N, ncol = 2*N)
    ```

3. **Phenotypic Value Calculation**:
    - For each plant, calculate its DGE and sum the IGEs from its neighbors.
    - Add environmental effects to obtain the total phenotypic value.

    ```r
    Valeur_G <- function(x, y, genotype, voisinage) {
        geno <- c()
        for (i in (x-1):(x+1)) {
            for (j in (y-1):(y+1)) {
                if (!(i == x && j == y)) {
                    geno <- c(geno, voisinage[i, j])
                }
            }
        }
        DGE <- genotype[voisinage[x, y], 1] + enviro[voisinage[x, y], 1]
        IGE <- sum(genotype[geno, 2] + enviro[geno, 2])
        return(DGE + IGE)
    }

    Pheno <- c()
    tri_geno <- c()
    cpt <- 0

    for (x in 5:(N+5)) {
        for (y in 5:(N+5)) {
            cpt <- cpt + 1
            Pheno[cpt] <- Valeur_G(x, y, genotype, voisinage)
            tri_geno[cpt] <- voisinage[x, y]
        }
    }
    ```

4. **Selection**:
    - Rank individuals by their phenotypic values.
    - Select the top proportion (\(p\)) of individuals.

    ```r
    N_sel <- round(cpt * input$p)
    geno_sel <- genotype[tri_geno[order(Pheno, decreasing = TRUE)][1:N_sel], ]
    ```

5. **Next Generation Prediction**:
    - Calculate the mean genetic values for the next generation based on the selected individuals.
    - Adjust the mean values considering the response to selection, including the correlation between DGE and IGE.

    ```r
    S <- c()
    S[1] <- mean(genotype[tri_geno[order(Pheno, decreasing = TRUE)][1:N_sel], 1])
    S[2] <- mean(genotype[tri_geno[order(Pheno, decreasing = TRUE)][1:N_sel], 2])

    R <- G %*% solve(P) %*% S
    mus <- mu
    mus[1] <- mu[1] + R[1]
    mus[2] <- mu[2] + R[2]

    new_geno <- mvrnorm(n = N^2, mu = mus, Sigma = G)
    new_voisinage <- matrix(sample(1:N^2, 4*N^2, replace = TRUE), nrow = 2*N, ncol = 2*N)

    new_Pheno <- c()
    new_tri_geno <- c()
    cpt <- 0

    for (x in 5:(N+5)) {
        for (y in 5:(N+5)) {
            cpt <- cpt + 1
            new_Pheno[cpt] <- Valeur_G(x, y, new_geno, new_voisinage)
            new_tri_geno[cpt] <- new_voisinage[x, y]
        }
    }
    ```

By understanding these steps and using the Shiny application, you will gain insights into the dynamics of genetic selection and the impact of social interactions on breeding outcomes.

#### Shiny Application Instructions
To use the Shiny application for this practical session, follow these steps:

1. Open the Shiny application in your web browser.
2. Set the number of individuals (N) using the slider.
3. Adjust the genetic variances for DGE and IGE, the genetic correlation, and the environmental variances using the provided inputs.
4. Set the selection pressure to determine the proportion of top-performing plants to be selected.
5. If desired, set a seed for reproducibility by checking the box and entering a seed value.
6. Click the "Run Simulation" button to start the simulation.
7. View the results in the "Graph" and "Summary" tabs, which show the genotype distribution, phenotypic distributions before and after selection, and a summary of the calculations.

#### Suggested Parameters
To help you understand the effects of selection, try using the following parameter values:

- **Number of individuals (N)**: 20
- **Genetic variance DGE (\(V_{DGE}\))**: 0.75
- **Genetic variance IGE (\(V_{IGE}\))**: 0.07
- **Genetic correlation DGE : IGE (\(r\))**: 0.5
- **Environmental variance DGE

 (\(V_{E_{DGE}}\))**: 0.7
- **Environmental variance IGE (\(V_{E_{IGE}}\))**: 0.07
- **Selection pressure (\(p\))**: 0.3

Observe how changes in these parameters affect the selection outcomes and the predicted genetic values of the next generation.